\section{Introduction}
\label{sec:intro}

\subsection{Motivation}
Machine-learning systems in scientific and safety-critical domains are almost
universally judged by a single scalar: accuracy, or one of its robust variants.
This is convenient, but it is also mathematically incomplete. A model may answer a
scientific question correctly and yet, under an adversarially crafted prompt,
silently violate the ethical requirements we care about---fabricating citations,
amplifying bias, leaking private information, or abandoning transparency---while
its answer remains nominally right. An accuracy-only report certifies such a model
as safe. The gap is not a matter of measurement noise: it is structural. A metric
that depends on the model only through its correctness behavior \emph{cannot}, as a
matter of logic, detect a failure that leaves correctness untouched.

This motivates a precise mathematical question. Adversarial-robustness theory has
produced rigorous, quantitative coefficients for the stability of
\emph{correctness} under perturbation---the CLEVER score built from local Lipschitz
constants \cite{Weng2018}, the certified radius of randomized smoothing
\cite{Cohen2019}, and the exact robust-error decomposition of TRADES
\cite{Zhang2019}. Alignment research has, in parallel, produced quantitative
measures of \emph{ethical} stability, chiefly the attack success rate against
standardized jailbreak benchmarks \cite{Chao2024,Zou2023}. These two literatures
have not been fused: to our knowledge no prior work combines a
certified-robustness-style correctness term with an alignment-stability term into a
single coefficient with provable properties, and none connects a training-time
ethical constraint to a lower bound on such a coefficient.

\subsection{The question}
We study the following informal hypothesis, which arises repeatedly in the
empirical literature on aligned models: \emph{models trained with explicit ethical
constraints are at least $0.15$ more robust than accuracy-only models under
adversarial prompts, and this improvement is captured by a composite metric.} We
ask whether this can be turned into a theorem. Concretely: is there a precisely
defined coefficient $\RC_\lambda$ with provable analytic properties, together with
explicit sufficient conditions under which a training-time ethical constraint
\emph{guarantees} a separation
\[
  \Delta \;=\; \E[\RC_\lambda^{\eth}] - \E[\RC_\lambda^{\acc}] \;\ge\; 0.15\,?
\]
And is such a composite \emph{necessary}, or does accuracy alone already suffice?

\subsection{Contributions}
We answer both questions affirmatively. We define the composite ethical-robustness
coefficient
\[
  \RC_\lambda(f,x,\varepsilon) \;=\; \lambda\,\rho_C(f,x,\varepsilon)
    + (1-\lambda)\,\rho_A(f,x,\varepsilon), \qquad \lambda\in[0,1],
\]
where $\rho_C,\rho_A\in[0,1]$ are the worst-case preservation ratios of scientific
correctness $C$ and ethical alignment $A$ over an adversarial perturbation set
$B(x,\varepsilon)$, and prove:

\begin{itemize}[leftmargin=*,itemsep=2pt,topsep=2pt]
  \item \textbf{Well-posedness (\Cref{thm:wellposed}).} $\RC_\lambda$ is bounded in
  $[0,1]$, non-increasing in $\varepsilon$, equal to $1$ at $\varepsilon=0$, and
  recovers the single-term correctness and alignment metrics at $\lambda=1$ and
  $\lambda=0$.
  \item \textbf{Exact decomposition (\Cref{thm:decomp}).} The deficit of each
  component obeys an exact additive identity $e_\rob^S=e_\nat^S+b^S$, the direct
  analogue of the TRADES identity $R_\rob=R_\nat+R_\mathrm{bdy}$ \cite{Zhang2019},
  yielding a normalized upper bound on $1-\E[\RC_\lambda]$.
  \item \textbf{Constraint $\Rightarrow$ separation (\Cref{thm:main}, central).} A
  constrained-Markov-decision-process alignment constraint on the ethical model,
  together with a bounded correctness cost $\delta_C$, forces
  $\Delta \ge (1-\lambda)\big[(1-d)-\bar a/a_{\min}\big]-\lambda\delta_C$, which
  meets the $0.15$ target on an explicit region.
  \item \textbf{Necessity of the composite (\Cref{thm:necessity}).} There exist two
  models indistinguishable to every correctness-only metric yet separated by
  $\RC_\lambda$; hence $\lambda<1$ is necessary to detect alignment collapse and
  $\lambda>0$ to detect correctness collapse.
  \item \textbf{Certified lower bounds (\Cref{thm:certified}).} $\RC_\lambda$
  admits attack-independent Lipschitz and randomized-smoothing certified lower
  bounds.
  \item \textbf{Admissible-weight region (\Cref{thm:lambda}).} The set of weights
  $\lambda$ achieving any target separation is characterized in closed form.
\end{itemize}

Although the correctness score $C$ is specialized in our motivating application to
correctness on astrophysics questions, every theorem below is stated for arbitrary
measurable scores $C,A\in[0,1]$ and is therefore domain-agnostic.

\subsection{Techniques}
The proofs are elementary but structured. Well-posedness and the decomposition
follow from the nesting property of perturbation sets and linearity of
expectation. The central separation theorem imports constrained-Markov-decision-
process feasibility \cite{Achiam2017} as the mechanism that turns an ethical
\emph{training constraint} into an alignment-robustness \emph{floor}, and combines
it with a clean-competence floor via two elementary bounds on the preservation
ratio. Necessity is proved by an explicit two-model hard-instance construction in
the style of \cite{Tsipras2018}. The certified bounds instantiate the Lipschitz
margin argument of \cite{Weng2018} and the Neyman--Pearson certificate of
\cite{Cohen2019}. Every symbolic identity and inequality is verified with a
computer-algebra system, and every quantitative claim is checked by a seeded
Monte-Carlo study; \Cref{sec:discussion} reports these.

\subsection{Organization}
\Cref{sec:prelim} fixes notation, states the definitions, and recalls the
prerequisite results. \Cref{sec:results} states and proves the six theorems, with
worked examples. \Cref{sec:discussion} reports the computational verification and
discusses implications, limitations, and open problems. \Cref{sec:conclusion}
concludes.
